---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Regular time series

If a time series is "regular", that is it has equal time steps between observations, and no missing observations, then its power spectrum can be estimated from its Fourier transform. 


## The Fourier Transform

```{r}
sunspot <- datasets::sunspot.month
plot(sunspot)
```

```{r}
library(PaleoSpec)
sp_sun <- SpecMTM(sunspot)
LPlot(sp_sun)
```


## The Multitaper Method



## To detrend or not {#sec-detrend}


### Why detrend?



```{r}
library(dplyr)
library(PaleoSpec)
library(ggplot2)

f1 <- 1/10
f2 <- 1/100
f3 <- 1/1000

tau <- 1e04

dat <- tibble(
  time = seq(0, tau, by = 1),
  y = cos(2*pi*f1*time) + cos(2*pi*f2*time) + cos(2*pi*f3*time) 
) %>% 
  mutate(y = y + rnorm(n(), 0, 0.1))

plot(dat$time, dat$y, type = "l")


sp1 <- SpecMTM(ts(dat$y, deltat = 1), detrend =  FALSE)
sp1_detr <- SpecMTM(ts(dat$y, deltat = 1), detrend =  TRUE)

#sp1 <- SpecACF(ts(dat$y, deltat = 1), bin.width = 1, detrend =  FALSE)


gg_spec(list(sp1, detrended = sp1_detr)) +
  geom_vline(xintercept = c(f1, f2, f3), lty = 2, colour =  "Red")



sp1_10k <- SpecACF(ts(dat$y, deltat = 1), bin.width = 1, detrend =  FALSE)
sp1_10k_detr <- SpecACF(ts(dat$y, deltat = 1), bin.width = 1, detrend =  TRUE)

plot(y~time, data = dat[1:500,],  type = "l")

sp1_0.5k <- SpecACF(ts(dat$y[1:500], deltat = 1), bin.width = 1, detrend =  FALSE)
sp1_0.5k_detr <- SpecACF(ts(dat$y[1:500], deltat = 1), bin.width = 1, detrend =  TRUE)


sp1_0.5k_detr_MTM <- SpecMTM(ts(dat$y[1:500], deltat = 1), detrend =  TRUE)


#sp1 <- SpecACF(ts(dat$y, deltat = 1), bin.width = 1, detrend =  FALSE)


gg_spec(list(
  raw_10k = sp1_10k, detrended_10K = sp1_10k_detr,
  raw_0.5k = sp1_0.5k, detrended_0.5K = sp1_0.5k_detr,
  detrended_0.5k_MTM = sp1_0.5k_detr_MTM
  )) +
  geom_vline(xintercept = c(f1, f2, f3), lty = 2, colour =  "Red")



```



- Even with detrend = FALSE, SpecMTM does some detrending - must be part of the tapering?



```{r}
library(dplyr)
library(tidyr)
library(PaleoSpec)
library(ggplot2)


N <- 1e03
n_rep <- 1000
alpha = 0.1
beta = 0

dat <- crossing(
  rep = 1:n_rep,
  t = 1:N
) #%>% 
  # group_by(rep) %>% 
  # mutate(
  #   y = PaleoSpec::SimPLS(N = N, beta = beta, alpha = alpha) 
  # )

# Get a matrix of simulated timeseries
ts_m <- replicate(n_rep, PaleoSpec::SimPLS(N = N, beta = beta, alpha = alpha))

mean_spec <- SpecACF(ts_m, bin.width = 1, detrend = FALSE)
mean_spec_detrended <- SpecACF(ts_m, bin.width = 1)

dat$y <- as.numeric(ts_m)

specs_mtm_detrend <- dat %>% 
  group_by(rep) %>% 
  do({
    Spec2DF(SpecMTM(ts(.$y), 
                    detrend = TRUE
                    )
            )
  })

specs_mtm_raw <- dat %>% 
  group_by(rep) %>% 
  do({
    Spec2DF(SpecMTM(ts(.$y), 
                    detrend = FALSE
                    )
            )
  })


mean_spec_MTM_raw <- specs_mtm_raw %>% 
  group_by(freq) %>% 
  summarise_if(is.numeric, mean)

mean_spec_MTM_detrended <- specs_mtm_detrend %>% 
  group_by(freq) %>% 
  summarise_if(is.numeric, mean)

gg_spec(list("Periodogram_raw_ts" = mean_spec,
             "Periodogram_detrended_ts" = mean_spec_detrended,
             MTM_detrended = mean_spec_MTM_detrended,
             MTM_raw_ts = mean_spec_MTM_raw)) #+
  #geom_point(aes(x = freq, y = spec, colour = spec_id))
```




## Errors

- Explain Gamma / Chi-Sq nature of the errors
- Degrees of freedom / shape of Gamma
- Effect of tapering and smoothing on error distribution
- Confidence intervals

