---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Regular time series

If a time series is "regular", that is it has equal time steps between observations, and no missing observations, then its power spectrum can be estimated from its Fourier transform. 


## The Fourier Transform

```{r}
sunspot <- datasets::sunspot.month
plot(sunspot)
```

```{r}
library(PaleoSpec)
sp_sun <- SpecMTM(sunspot)
LPlot(sp_sun)
```


## The Multitaper Method



## To detrend or not 

- Even with detrend = FALSE, SpecMTM does some detrending - must be part of the tapering?


```{r}
library(dplyr)
library(tidyr)
library(PaleoSpec)
library(ggplot2)


N <- 1e03
n_rep <- 100
alpha = 0.1
beta = 0

dat <- crossing(
  rep = 1:n_rep,
  t = 1:N
) %>% 
  group_by(rep) %>% 
  mutate(
    y = PaleoSpec::SimPLS(N = N, beta = beta, alpha = alpha) + 0.1 * t
  )


specs <- dat %>% 
  group_by(rep) %>% 
  do({
    Spec2DF(SpecMTM(ts(.$y), 
                    detrend = FALSE,
                    #centre = "arithMean"
                    #K = 0, nw = 1
                    )
            )
  })

specs <- dat %>% 
  group_by(rep) %>% 
  do({
    Spec2DF(SpecACF(ts(.$y), bin.width = 1,
                    detrend = FALSE, 
                    #K = 0, nw = 1
                    )
            )
  })



mean_spec <- specs %>% 
  group_by(freq) %>% 
  summarise_if(is.numeric, mean)


mean_spec %>% 
  as_spec_df() %>% 
  gg_spec() +
  #geom_point() +
  geom_abline(intercept = log10(alpha), slope = -beta, colour = "red") #+
  #expand_limits(y = 100)
```

```{r}
library(multitaper)
x <- 1:101
y <- 1.0 + 0.5*(x) + rnorm(n=101,mean=0,sd=2)
vars <- multitaperTrend(xd=y, B=0.05, deltat=1.0, t.in=x)
plot(x,y,type="l")
lines(x,vars[[1]]+vars[[2]]*vars[[3]],type="l",col="red")
```




## Errors

- Explain Gamma / Chi-Sq nature of the errors
- Degrees of freedom / shape of Gamma
- Effect of tapering and smoothing on error distribution
- Confidence intervals

