---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Variance from a power spectrum

The integral of the full power spectrum is equal to the variance of the whole time series. We can calculate this by summing the spectral estimates and multiplying by delta_f. 


```{r setup}
library(PaleoSpec)
library(ggplot2)
```

```{r}
set.seed(20240228)
N <- 1e03
delta_t = 1

ts1 <- ts(rnorm(N), deltat = delta_t) * 5
sp1 <- SpecMTM(ts1, detrend = TRUE)
```

```{r}
gg_spec(sp1) +
  scale_x_continuous() 
```

Note that SpecMTM returns only the positive frequencies, therefore to get the total variance we must multiply by a factor of 2.

```{r}
var(ts1)

2 * sum(sp1$spec) * diff(sp1$freq)[1]
```

Why are these not identical?

## Variance by timescale

As we simulated a "white" time series, the total variance of the time series is spread equally across all frequencies. Therefore the integrals of the lower and upper halves of the power spectrum are expected to be equal to each other and sum to the variance of the whole time series.

PaleoSpec has a function `GetVarFromSpectra()` which we will use to integrate sections of the power spectrum

```{r}

f_range <- c(1/N, 1/2)
f_range_low <- c(1/N, mean(f_range))
f_range_high <- c(mean(f_range), 1/2)

PaleoSpec::GetVarFromSpectra(sp1, f_range)

PaleoSpec::GetVarFromSpectra(sp1, f_range_low)
PaleoSpec::GetVarFromSpectra(sp1, f_range_high)


PaleoSpec::GetVarFromSpectra(sp1, f_range_low)$var +
PaleoSpec::GetVarFromSpectra(sp1, f_range_high)$var

```


In contrast, if we simulated a time series where the variance increases with timescale, integrating different parts of its power spectrum will yield different variances.


```{r}
ts2 <- ts(PaleoSpec::SimPLS(N, beta = 1, alpha = 0.1))
plot(ts2)

sp2 <- SpecMTM(ts2)

gg_spec(sp2)  +
  scale_x_continuous()+
  geom_vline(xintercept = mean(f_range), colour = "red")


var_low_2 <- PaleoSpec::GetVarFromSpectra(sp2, f_range_low)
var_high_2 <- PaleoSpec::GetVarFromSpectra(sp2, f_range_high)

var_low_2
var_high_2
```

The sum of the two variances calculated by integrating the two halves of the power spectrum should still be equal to the total variance of the time series.

```{r}
var_low_2$var + var_high_2$var
var(ts2)
```

When a time series has increasing power at lower frequencies, the variance estimated from the power spectrum will be lower than the true variance, because the detrending of the time series prior to estimating the power spectrum removes power from the lowest frequencies. See @sec-detrend .


