# Variance from a power spectrum

The integral of the full power spectrum is equal to the variance of the whole time series. We can calculate this by summing the spectral estimates and multiplying by delta_f. 


```{r setup}
library(PaleoSpec)
library(ggplot2)
```

```{r}
set.seed(20240228)
N <- 1e03
delta_t = 1

ts1 <- ts(rnorm(N), deltat = delta_t) * 5
sp1 <- SpecMTM(ts1, detrend = TRUE)
```

```{r}
gg_spec(sp1) +
  scale_x_continuous() 
```

Note that SpecMTM returns only the positive frequencies, therefore to get the total variance we must multiply by a factor of 2.

```{r}
var(ts1)

2 * sum(sp1$spec) * diff(sp1$freq)[1]
```

Why are these not identical?

## Variance by timescale

As we simulated a "white" time series, the total variance of the time series is spread equally across all frequencies. Therefore the integrals of the lower and upper halves of the power spectrum are expected to be equal to each other and sum to the variance of the whole time series.

PaleoSpec has a function `GetVarFromSpectra()` which we will use to integrate sections of the power spectrum

```{r}

f_range <- c(1/N, 1/2)
f_range_low <- c(1/N, mean(f_range))
f_range_high <- c(mean(f_range), 1/2)

PaleoSpec::GetVarFromSpectra(sp1, f_range)

PaleoSpec::GetVarFromSpectra(sp1, f_range_low)
PaleoSpec::GetVarFromSpectra(sp1, f_range_high)


PaleoSpec::GetVarFromSpectra(sp1, f_range_low)$var +
PaleoSpec::GetVarFromSpectra(sp1, f_range_high)$var

```
